---
title: "Statistical_Analysis.Rmd"
author: "Daniel Brock"
date: "8/2/2023"
output: html_document
---

# 0. Importing Packages and setting working directory

```{r setup, include=FALSE}
# Packages 
library(plyr)
library(tidyverse)
library(finalfit)
library(broom)
library(ggfortify)
library(ggthemes)
library(patchwork)
library(magrittr)
library(forcats)
library(rstatix)
library(rcompanion)
library(survival)
library(survminer)
library(readxl)
library(writexl)
library(pheatmap)
library(zoo)
library(sjPlot)
library(DHARMa)

# Setting working directory
cwd <- getwd()
```

# 1. Importing merged trauma dataset from python export

```{r}
# Importing datasets
trauma <- read_csv(file = paste0(cwd, "/TQP_Processed/trauma_merged.csv"))
comor <- read_csv(file = paste0(cwd, "/TQP_Processed/comorbidities_merged.csv"))
adverse <- read_csv(file = paste0(cwd, "/TQP_Processed/adverse_events_merged.csv"))
ais <- read_csv(file = paste0(cwd, "/TQP_Processed/ais_merged.csv"))

# Calculating total NTDB cases (from python file)
total_cases <- 997970 + 1043736 + 1097190 + 1133053 + 1209097 + 1232956 + 1300735
print(total_cases)
```

## 1.1 Viewing the data and dropping important NA rows

```{r}
# Replacing quote marks in ecodes
trauma$ECODE_DESC <- gsub(pattern = "\"", replacement = "", x = trauma$ECODE_DESC)

# Merging trauma with comorbidities
trauma <- merge(trauma, comor, on = "INC_KEY")

# Merging trauma with adverse events
trauma <- merge(trauma, adverse, on = "INC_KEY")

# Glimpsing
trauma_glimpse <- ff_glimpse(trauma)
trauma_glimpse_continuous <- trauma_glimpse$Continuous
trauma_glimpse_categorical <- trauma_glimpse$Categorical
```

## 1.2 Refactoring data for table 1

```{r}
# Refactoring Sex
trauma <- trauma %>% filter(SEX != 3) %>% mutate(
  sex.factor = SEX %>% 
    factor() %>% 
    fct_recode("Male" = "1", "Female" = "2") %>% 
    ff_label("Sex")
)

# Refactoring Age
trauma <- trauma %>% 
  mutate(
    age.factor = 
      AGEYEARS %>%
      cut(breaks = c(0,10,20,30,40,50,60,70,100), include.lowest = TRUE, right = FALSE) %>% 
      fct_recode(
        "0-9"   =  "[0,10)",
        "10-19" = "[10,20)",
        "20-29" = "[20,30)",
        "30-39" = "[30,40)",
        "40-49" = "[40,50)",
        "50-59" = "[50,60)",
        "60-69" = "[60,70)",
        "70+"   = "[70,100]",
      ) %>% 
      ff_label("Age (years)")
    )

# Refactoring Race 
trauma <- trauma %>% mutate(
  race.factor = case_when(
    ASIAN == 1 ~ "Asian",
    PACIFICISLANDER == 1 ~ "Pacific Islander",
    RACEOTHER == 1 ~ "Other",
    AMERICANINDIAN == 1 ~ "Native American",
    BLACK == 1 ~ "Black",
    WHITE == 1 ~ "White",
    RACE_NA == 1 ~ "NA",
    RACE_UK == 1 ~ "NA"
  )
)
trauma <- trauma %>% mutate(
  race.factor = na_if(race.factor, "NA") %>% 
    factor(levels = c("White", "Black", "Asian", "Native American", "Pacific Islander", "Other")) %>% 
    ff_label("Race")
)

# Refactoring Ethnicity
trauma <- trauma %>% mutate(
  ethnicity.factor = ETHNICITY %>% 
    factor() %>% 
    fct_recode("Hispanic" = "1", "Not Hispanic" = "2") %>% 
    ff_label("Ethnicity")
)

# Refactoring Helmet Wearing
trauma <- trauma %>% mutate(
  helmet.factor = case_when(
    PROTDEV_HELMET == 1 & PROTDEV_UK != 1 ~ "Helmet",
    PROTDEV_HELMET == 0 & PROTDEV_UK != 1 ~ "No Helmet",
    PROTDEV_UK == 1 ~ "NA"
  )
)
trauma <- trauma %>% mutate(
  helmet.factor = na_if(helmet.factor, "NA") %>% 
    factor(levels = c("Helmet", "No Helmet")) %>% 
    ff_label("Helmet Use")
)

# Refactoring hypotension
trauma <- trauma %>% mutate(
  SBP = SBP %>% ff_label("SBP mmHg")
)
trauma <- trauma %>% mutate(
  sbp.hypo = case_when(
    SBP < 90 ~ "Hypotensive",
    SBP >= 90 ~ "Not Hypotensive"
  ) %>% 
    ff_label("SBP mmHg <90")
)

# Refactoring TOTALGCS
trauma <- trauma %>% mutate(
  TOTALGCS = TOTALGCS %>% ff_label("Total GCS")
)


# Multiple drug use
trauma$total.drugs <- trauma$DRGSCR_AMPHETAMINE + trauma$DRGSCR_BARBITURATE + trauma$DRGSCR_BENZODIAZEPINES + trauma$DRGSCR_COCAINE + trauma$DRGSCR_METHAMPHETAMINE + trauma$DRGSCR_ECSTASY + trauma$DRGSCR_METHADONE + trauma$DRGSCR_OPIOID + trauma$DRGSCR_OXYCODONE + trauma$DRGSCR_PHENCYCLIDINE + trauma$DRGSCR_TRICYCLICDEPRESS + trauma$DRGSCR_CANNABINOID + trauma$DRGSCR_OTHER

# Alcohol use categories
trauma <- trauma %>% mutate(
  alcohol.use = case_when(
    is.na(ALCOHOLSCREENRESULT) ~ "Not Screened",
    ALCOHOLSCREENRESULT < 0.02 & total.drugs == 0 ~ "Sober",
    ALCOHOLSCREENRESULT >= 0.02 & ALCOHOLSCREENRESULT < 0.08 & total.drugs == 0 ~ "Impaired",
    ALCOHOLSCREENRESULT >= 0.08 & total.drugs == 0 ~ "Intoxicated",
    ALCOHOLSCREENRESULT >= 0.08 & total.drugs >= 1 ~ "Multi-Drug"
  ) %>% 
    factor(levels = c("Not Screened", "Sober", "Impaired", "Intoxicated", "Multi-Drug")) %>% 
    ff_label("Alcohol")
)

# Refactoring Drug use
trauma <- trauma %>%
  mutate(
    drug.use = case_when(
      DRGSCR_NONE == 1 & total.drugs < 1 & ALCOHOLSCREENRESULT < 0.02 ~ "None",
      DRGSCR_AMPHETAMINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Amphetamine",
      DRGSCR_BARBITURATE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Barbiturate",
      DRGSCR_BENZODIAZEPINES == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Benzodiazepines",
      DRGSCR_COCAINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Cocaine",
      DRGSCR_METHAMPHETAMINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Methamphetamine",
      DRGSCR_ECSTASY == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Ecstasy",
      DRGSCR_METHADONE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Methadone",
      DRGSCR_OPIOID == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Opioids",
      DRGSCR_OXYCODONE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Oxycodone",
      DRGSCR_PHENCYCLIDINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Phencyclidine",
      DRGSCR_TRICYCLICDEPRESS == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Tricyclic Antidepressants",
      DRGSCR_CANNABINOID == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Cannabis",
      DRGSCR_OTHER == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Other",
      total.drugs >= 2 & ALCOHOLSCREEN < 0.02 ~ "Multiple Drugs",
      total.drugs >= 1 & ALCOHOLSCREENRESULT >= 0.02 ~ "Drugs and Alcohol",
      TRUE ~ NA_character_  # Add a default value
    ) %>% 
      factor(levels = c("None", "Cannabis", "Amphetamine", "Opioids", "Cocaine", "Benzodiazepines", "Methamphetamine", "Oxycodone", "Barbiturate", "Phencyclidine", "Tricyclic Antidepressants", "Methadone", "Ecstasy", "Drugs and Alcohol", "Other")) %>% 
      ff_label("Drugs")
  )

# Refactoring length of hospital stay
trauma <- trauma %>% mutate(
  FINALDISCHARGEDAYS = FINALDISCHARGEDAYS %>% ff_label("Length of Hospital Stay (days)")
)

# Refactoring length of ICU stay
trauma <- trauma %>% mutate(
  TOTALICULOS = TOTALICULOS %>% ff_label("Length of ICU Stay (days)")
)

# Refactoring ISS (injury severity score)
trauma <- trauma %>% mutate(
  ISS = ISS %>% ff_label("Injury Severity Score")
)

trauma <- trauma %>% mutate(
  ISS.factor = ISS %>% 
    cut(breaks = c(0,3,8,15,24,75), include.lowest = TRUE, right = TRUE) %>% 
    fct_recode(
      "Minor (0-3)" = "[0,3]",
      "Moderate (4-8)" = "(3,8]",
      "Serious (9-15)" = "(8,15]",
      "Severe (16-24)" = "(15,24]",
      "Critical (>25)" = "(24,75]"
    ) %>% 
    ff_label("ISS Categories")
)

# Refactoring Comorbidities
trauma <- trauma %>% mutate(
  CC_ALCOHOLISM = CC_ALCOHOLISM %>% ff_label("Alcoholism"),
  CC_SMOKING = CC_SMOKING %>% ff_label("Smoker"),
  CC_SUBSTANCEABUSE = CC_SUBSTANCEABUSE %>% ff_label("Substance Abuse Disorder"),
  CC_CHF = CC_CHF %>% ff_label("Congestive Heart Failure"),
  CC_RENAL = CC_RENAL %>% ff_label("End Stage Renal Disease"),
  CC_DIABETES = CC_DIABETES %>% ff_label("Diabetes"),
  CC_HYPERTENSION = CC_HYPERTENSION %>% ff_label("Hypertension"),
  CC_COPD = CC_COPD %>% ff_label("COPD"),
  CC_BLEEDING = CC_BLEEDING %>% ff_label("Bleeding Disorder"),
  CC_ANTICOAGULANT = CC_ANTICOAGULANT %>% ff_label("Anticoagulant Therapy"),
  CC_CIRRHOSIS = CC_CIRRHOSIS %>% ff_label("Cirrosis"),
  CC_CVA = CC_CVA %>% ff_label("Cerebrovascular Accident"),
  CC_MENTALPERSONALITY = CC_MENTALPERSONALITY %>% ff_label("Mental/Personality Disorder")
)

# Refactoring Adverse Events
trauma <- trauma %>% mutate(
  HC_ALCOHOLWITHDRAWAL = HC_ALCOHOLWITHDRAWAL %>% ff_label("Alcohol Withdrawal Syndrome"),
  HC_CARDARREST = HC_CARDARREST %>% ff_label("Cardiac Arrest"),
  HC_DVTHROMBOSIS = HC_DVTHROMBOSIS %>% ff_label("Deep Vein Thrombosis"),
  HC_EMBOLISM = HC_EMBOLISM %>% ff_label("Pulmonary Embolism"),
  HC_KIDNEY = HC_KIDNEY %>% ff_label("Acute Kidney Injury"),
  HC_RESPIRATORY = HC_RESPIRATORY %>% ff_label("Acute Respiratory Distress Syndrome"),
  HC_SEPSIS = HC_SEPSIS %>% ff_label("Sepsis"),
  HC_STROKECVA = HC_STROKECVA %>% ff_label("Stroke / CVA"),
  HC_VAPNEUMONIA = HC_VAPNEUMONIA %>% ff_label("Hosipital-acquired Pneumonia"),
  HC_INTUBATION = HC_INTUBATION %>% ff_label("Unplanned Intubation"),
  HC_UNPLANNEDICU = HC_UNPLANNEDICU %>% ff_label("Unplanned Admission to ICU"),
  HC_RETURNOR = HC_RETURNOR %>% ff_label("Unplanned Return to OR")
)

# Refactoring mortality using pulserate - this is not mortality!
#trauma <- trauma %>% mutate(
#  mortality.factor = case_when(
#    PULSERATE == 0 ~ "Death",
#    PULSERATE != 0 ~ "Alive"
#  ) %>% 
#    factor(levels = c("Death", "Alive")) %>% 
#    ff_label("Mortality")
#)

# Correctly refactoring mortality using ED and Hospital discharge disposition
alive_dispositions <- c(1, 2, 3, 4, 6, 7, 8, 9, 10, 11, 12, 13, 14)
trauma <- trauma %>% mutate(
  mortality.total = case_when(
    EDDISCHARGEDISPOSITION == 5 | HOSPDISCHARGEDISPOSITION == 5 ~ "Deceased",
    EDDISCHARGEDISPOSITION %in% alive_dispositions | HOSPDISCHARGEDISPOSITION %in% alive_dispositions ~ "Alive"
  ) %>% 
    factor(levels = c("Deceased", "Alive")) %>% 
    ff_label("Total Mortality")
)
trauma <- trauma %>% mutate(
  mortality.ED = case_when(
    EDDISCHARGEDISPOSITION == 5 ~ "Deceased",
    EDDISCHARGEDISPOSITION %in% alive_dispositions ~ "Alive"
  ) %>% 
    factor(levels = c("Deceased", "Alive")) %>% 
    ff_label("ED Mortality")
)
trauma <- trauma %>% mutate(
  mortality.hospital = case_when(
    HOSPDISCHARGEDISPOSITION == 5 ~ "Deceased",
    HOSPDISCHARGEDISPOSITION %in% alive_dispositions ~ "Alive"
  ) %>% 
    factor(levels = c("Deceased", "Alive")) %>% 
    ff_label("Hospital Mortality")
)

# Refactoring alcohol withdrawal syndrome
trauma <- trauma %>% mutate(
  alc.withdraw.factor = HC_ALCOHOLWITHDRAWAL %>% 
    factor() %>% 
    fct_recode("Positive" = "1", "None" = "0") %>% 
    ff_label("Alcohol Withdrawal Syndrome")
)

# Refactoring Year
trauma <- trauma %>% mutate(
  year.factor = Year %>% 
    factor() %>% 
    ff_label("Year")
)
```

## 1.3 Filtering for NAs in important variables

```{r}
# Filtering for NAs in selected columns
trauma_filt <- trauma %>% filter_at(vars(sex.factor, age.factor, race.factor, ethnicity.factor, helmet.factor), all_vars(!is.na(.)))
#initial filter included: sex.factor, age.factor, race.factor, ethnicity.factor, helmet.factor, SBP, sbp.hypo, TOTALGCS, FINALDISCHARGEDAYS, ISS, ISS.factor, CC_ALCOHOLISM, CC_SMOKING, CC_SUBSTANCEABUSE, CC_CHF, CC_RENAL, CC_DIABETES, CC_HYPERTENSION, CC_COPD, alc.withdraw.factor, mortality.factor

# Seeing the total number of BAC screens for people with and without other drug use
bac_screens <- trauma_filt %>% mutate(
  alcohol.use.total = case_when(
    is.na(ALCOHOLSCREENRESULT) ~ "Not Screened",
    ALCOHOLSCREENRESULT < 0.02 ~ "Sober",
    ALCOHOLSCREENRESULT >= 0.02 & ALCOHOLSCREENRESULT < 0.08 ~ "Impaired",
    ALCOHOLSCREENRESULT >= 0.08 ~ "Intoxicated"
  ),
  alcohol.use.clean = case_when(
    is.na(ALCOHOLSCREENRESULT) ~ "Not Screened",
    ALCOHOLSCREENRESULT < 0.02 & total.drugs == 0 ~ "Sober",
    ALCOHOLSCREENRESULT >= 0.02 & ALCOHOLSCREENRESULT < 0.08 & total.drugs == 0 ~ "Impaired",
    ALCOHOLSCREENRESULT >= 0.08 & total.drugs == 0 ~ "Intoxicated",
    ALCOHOLSCREENRESULT >= 0.08 & total.drugs >= 0 ~ "Multi-Drug"
  )
)
bac_screens_total <- sum(!is.na(bac_screens$alcohol.use.total))
bac_screens_clean <- sum(!is.na(bac_screens$alcohol.use.clean))
print(paste0("Number of total BAC screens in patients with known info: ", bac_screens_total))
print(paste0("Number of filtered alcohol/multi-drug BAC screens in patients with known info: ", bac_screens_clean))

# Filtering for only patients with BAC screens with NO other drug use (clean)
#keys_to_keep <- bac_screens[!is.na(bac_screens$alcohol.use.clean), ] %>% pull(INC_KEY)
#trauma_filt <- trauma_filt %>% filter(INC_KEY %in% keys_to_keep)
```

```{r}
# Optional export to excel
#write_xlsx(trauma_filt, path = paste0(cwd, "/TQP_Processed/trauma_filtered.xlsx"))
```



# 2. Demographics table for alcohol use

```{r}
# Table on helmet use
# Explanatory & confounding variables
explanatory <- c("alcohol.use", "sex.factor", "age.factor", "race.factor", "ethnicity.factor")

# Dependent variable of interest
dependent <- "helmet.factor"

table1 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE,
                     add_col_totals = TRUE)

# Optional export to excel
#write_xlsx(x = table1, path = "tables/table1_demographics.xlsx")
```

```{r}
# Table on alcohol use groups
# Explanatory & confounding variables
explanatory <- c("helmet.factor", "sex.factor", "age.factor", "race.factor", "ethnicity.factor")

# Dependent variable of interest
dependent <- "alcohol.use"

table1.1 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE)
```


```{r}
# Pairwise comparison stats with row wise fisher's exact test for Alcohol use
rwf1 <- table(trauma_filt$alcohol.use, trauma_filt$helmet.factor) %>% as.data.frame.matrix()
rwf1 <- rwf1[ , c(2,1)]

pairwiseNominalMatrix(x = table(trauma_filt$alcohol.use, trauma_filt$helmet.factor), compare = "row", fisher = TRUE, chisq = FALSE)

rwf2 <- pairwise_fisher_test(rwf1, p.adjust.method = "fdr", detailed = TRUE)
chisq.test(rwf1)
```

```{r}
# Pairwise wilcox test on age
pairwise.wilcox.test(trauma_filt$AGEYEARS, trauma_filt$alcohol.use, p.adjust.method = "fdr")

chisq.test(table(trauma_filt$age.factor, trauma_filt$helmet.factor))
```

```{r}
# Pairwise comparison stats with sex
rwf1 <- table(trauma_filt$sex.factor, trauma_filt$helmet.factor) %>% as.data.frame.matrix()
rwf1 <- rwf1[ , c(2,1)]

rwf2 <- pairwise_fisher_test(rwf1, p.adjust.method = "fdr", detailed = TRUE)
chisq.test(rwf1)
```

```{r}
# Pairwise comparison stats with race
rwf1 <- table(trauma_filt$race.factor, trauma_filt$helmet.factor) %>% as.data.frame.matrix()
rwf1 <- rwf1[ , c(2,1)]

rwf2 <- pairwise_fisher_test(rwf1, p.adjust.method = "fdr", detailed = TRUE)
chisq.test(rwf1)
```

```{r}
# Pairwise comparison stats with ethnicity
rwf1 <- table(trauma_filt$ethnicity.factor, trauma_filt$helmet.factor) %>% as.data.frame.matrix()
rwf1 <- rwf1[ , c(2,1)]

rwf2 <- pairwise_fisher_test(rwf1, p.adjust.method = "fdr", detailed = TRUE)
chisq.test(rwf1)
```





# 3. Clinical presentation and alcohol use groups

```{r}
# Explanatory & confounding variables
explanatory <- c("mortality.total", "mortality.ED", "mortality.hospital", "FINALDISCHARGEDAYS", "TOTALICULOS", "ISS", "ISS.factor", "TOTALGCS", "SBP", "sbp.hypo", "HC_ALCOHOLWITHDRAWAL", "HC_CARDARREST", "HC_DVTHROMBOSIS", "HC_EMBOLISM", "HC_KIDNEY", "HC_RESPIRATORY", "HC_SEPSIS", "HC_STROKECVA", "HC_VAPNEUMONIA", "HC_INTUBATION", "HC_UNPLANNEDICU", "HC_RETURNOR")

# Dependent variable of interest
dependent <- "alcohol.use"  #helmet.factor

table2 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE,
                     add_row_totals = FALSE)

# Optional export to excel
#write_csv(x = table2, path = paste0(cwd, "/tables/table2_clinical_presentation.csv"))
```

```{r}
# Explanatory & confounding variables
explanatory <- c("mortality.total", "mortality.ED", "mortality.hospital", "FINALDISCHARGEDAYS", "TOTALICULOS", "ISS", "ISS.factor", "TOTALGCS", "SBP", "sbp.hypo", "HC_ALCOHOLWITHDRAWAL", "HC_CARDARREST", "HC_DVTHROMBOSIS", "HC_EMBOLISM", "HC_KIDNEY", "HC_RESPIRATORY", "HC_SEPSIS", "HC_STROKECVA", "HC_VAPNEUMONIA", "HC_INTUBATION", "HC_UNPLANNEDICU", "HC_RETURNOR")

# Dependent variable of interest
dependent <- "helmet.factor"  #helmet.factor or alcohol.use

table2.1 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE,
                     add_col_totals = TRUE)

# Optional export to excel
#write_xlsx(x = table2.1, path = "tables/table2_clinical_presentation.xlsx")
```

```{r}
# Mortality rates in helmet groups

# Total mortality
rwf1 <- table(trauma_filt$mortality.total, trauma_filt$helmet.factor)
rwf1 <- rwf1[ , c(2,1)]
fisher.test(rwf1)

# ED mortality
rwf1 <- table(trauma_filt$mortality.ED, trauma_filt$helmet.factor)
rwf1 <- rwf1[ , c(2,1)]
fisher.test(rwf1)

# Hospital mortality
rwf1 <- table(trauma_filt$mortality.hospital, trauma_filt$helmet.factor)
rwf1 <- rwf1[ , c(2,1)]
fisher.test(rwf1)
```

```{r}
# Length of hospital stay
wilcox.test(x = trauma_filt$FINALDISCHARGEDAYS, g = trauma_filt$helmet.factor)

# Length of ICU stay
wilcox.test(x = trauma_filt$TOTALICULOS, g = trauma_filt$helmet.factor)
```


```{r}
# ISS
wilcox.test(x = trauma_filt$ISS, g = trauma_filt$helmet.factor)

# Pairwise fisher test for ISS categories
rwf1 <- table(trauma_filt$ISS.factor, trauma_filt$helmet.factor)
rwf1 <- rwf1[, c(2,1)]
rwf2 <- row_wise_fisher_test(xtab = rwf1, p.adjust.method = "fdr", detailed = TRUE)
```

```{r}
# Total glasgow coma scale
wilcox.test(TOTALGCS ~ helmet.factor, data = trauma_filt)
```

```{r}
# Clinical outcomes
clin_out <- function(tab, outcome, flip = FALSE) {
  if (flip == TRUE) {
    tab <- tab[, c(2,1)]
  }
  fish <- fisher.test(tab)
  print(outcome)
  df <- data.frame("clinical_outcome" = outcome, "p_value" = fish$p.value, "OR" = fish$estimate, "conf_low" = fish$conf.int[1], "conf_high" = fish$conf.int[2])
  rownames(df) <- NULL
  return(df)
}

# hypotensive - change to tab switch 
c1 <- clin_out(tab = table(trauma_filt$sbp.hypo, trauma_filt$helmet.factor), outcome = "Hypotensive", flip = TRUE)

# alcohol withdrawal syndrome
c2 <- clin_out(tab = table(trauma_filt$alc.withdraw.factor, trauma_filt$helmet.factor), outcome = "Alcohol Withdrawal Sydrome")

# cardiac arrest
c3 <- clin_out(tab = table(trauma_filt$HC_CARDARREST, trauma_filt$helmet.factor), outcome = "Cardiac Arrest")

# DVT
c4 <- clin_out(tab = table(trauma_filt$HC_DVTHROMBOSIS, trauma_filt$helmet.factor), outcome = "DVT")

# Pulmonary embolism
c5 <- clin_out(tab = table(trauma_filt$HC_EMBOLISM, trauma_filt$helmet.factor), outcome = "Pulmonary Embolism")

# AKI
c6 <- clin_out(tab = table(trauma_filt$HC_KIDNEY, trauma_filt$helmet.factor), outcome = "Acute Kidney Injury")

# ARDS
c7 <- clin_out(tab = table(trauma_filt$HC_RESPIRATORY, trauma_filt$helmet.factor), outcome = "Acute Respiratory Distress Syndrome")

# Sepsis
c8 <- clin_out(tab = table(trauma_filt$HC_SEPSIS, trauma_filt$helmet.factor), outcome = "Sepsis")

# Stroke / CVA
c9 <- clin_out(tab = table(trauma_filt$HC_STROKECVA, trauma_filt$helmet.factor), outcome = "Stroke / CVA")

# Hospital-aquired pneumonia
c10 <- clin_out(tab = table(trauma_filt$HC_VAPNEUMONIA, trauma_filt$helmet.factor), outcome = "Hospital-aquired pneumonia")

# Unplanned intubation
c11 <- clin_out(tab = table(trauma_filt$HC_INTUBATION, trauma_filt$helmet.factor), outcome = "Unplanned Intubation")

# Unplanned admission to ICU
c12 <- clin_out(tab = table(trauma_filt$HC_UNPLANNEDICU, trauma_filt$helmet.factor), outcome = "Unplanned admission to ICU")

# Unplanned return to OR
c13 <- clin_out(tab = table(trauma_filt$HC_RETURNOR, trauma_filt$helmet.factor), outcome = "Unplanned admission to OR")

# Merging
c <- dplyr::bind_rows(list(c1, c2, c3, c4, c5, c6, c7, c8, c9, c10, c11, c12, c13))
c$OR_CI <- paste0(round(c$OR, 2), " [", round(c$conf_low, 2), "-", round(c$conf_high, 2), "]")

# Exporting to excel
#write_xlsx(x = c, path = "tables/adverse_outcomes_fisher.xlsx")
```





# 4. Comorbidities and alcohol use groups

```{r}
# Explanatory & confounding variables
explanatory <- c("CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE", "CC_MENTALPERSONALITY", "CC_CHF", "CC_RENAL", "CC_DIABETES", "CC_HYPERTENSION", "CC_COPD", "CC_BLEEDING", "CC_ANTICOAGULANT", "CC_CIRRHOSIS", "CC_CVA")

# Dependent variable of interest
dependent <- "alcohol.use"  #helmet.factor or alcohol.use

table3 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE,
                     add_row_totals = FALSE)

# Optional export to excel
#write_xlsx(x = table3, path = "tables/table3_comorbidities.xlsx")
```

```{r}
# Explanatory & confounding variables
explanatory <- c("CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE", "CC_MENTALPERSONALITY", "CC_CHF", "CC_RENAL", "CC_DIABETES", "CC_HYPERTENSION", "CC_COPD", "CC_BLEEDING", "CC_ANTICOAGULANT", "CC_CIRRHOSIS", "CC_CVA")

# Dependent variable of interest
dependent <- "helmet.factor"  #helmet.factor or alcohol.use

table3.1 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE,
                     add_col_totals = TRUE)

# Optional export to excel
#write_xlsx(x = table3.1, path = "tables/table3_comorbidities.xlsx")
```

```{r}
# Comorbidities fishers

# alcohol withdrawal syndrome
d1 <- clin_out(tab = table(trauma_filt$CC_ALCOHOLISM, trauma_filt$helmet.factor), outcome = "Alcohol Use Disorder")

# smoking
d2 <- clin_out(tab = table(trauma_filt$CC_SMOKING, trauma_filt$helmet.factor), outcome = "Tobacco Use Disorder")

# substance use
d3 <- clin_out(tab = table(trauma_filt$CC_SUBSTANCEABUSE, trauma_filt$helmet.factor), outcome = "Substance Use Disorder")

# mental/personality disorder
d4 <- clin_out(tab = table(trauma_filt$CC_MENTALPERSONALITY, trauma_filt$helmet.factor), outcome = "Mental/Personality Disorder")

# congestive heart failure
d5 <- clin_out(tab = table(trauma_filt$CC_CHF, trauma_filt$helmet.factor), outcome = "Congestive heart failure")

# end stage renal disease
d6 <- clin_out(tab = table(trauma_filt$CC_RENAL, trauma_filt$helmet.factor), outcome = "End stage renal disease")

# diabetes 
d7 <- clin_out(tab = table(trauma_filt$CC_DIABETES, trauma_filt$helmet.factor), outcome = "Diabetes")

# hypertension
d8 <- clin_out(tab = table(trauma_filt$CC_HYPERTENSION, trauma_filt$helmet.factor), outcome = "Hypertension")

# COPD
d9 <- clin_out(tab = table(trauma_filt$CC_COPD, trauma_filt$helmet.factor), outcome = "COPD")

# bleeding disorder
d10 <- clin_out(tab = table(trauma_filt$CC_BLEEDING, trauma_filt$helmet.factor), outcome = "Bleeding disorder")

# anticoagulant therapy
d11 <- clin_out(tab = table(trauma_filt$CC_ANTICOAGULANT, trauma_filt$helmet.factor), outcome = "Anticoagulant therapy")

# Cirrhosis
d12 <- clin_out(tab = table(trauma_filt$CC_CIRRHOSIS, trauma_filt$helmet.factor), outcome = "Cirrhosis")

# Cerebrovascular attack
d13 <- clin_out(tab = table(trauma_filt$CC_CVA, trauma_filt$helmet.factor), outcome = "Cerebrovascular attack")

# Merging
d <- dplyr::bind_rows(list(d1, d2, d3, d4, d5, d6, d7, d8, d9, d10, d11, d12, d13))
d$OR_CI <- paste0(round(d$OR, 2), " [", round(d$conf_low, 2), "-", round(d$conf_high, 2), "]")

# Exporting to excel
#write_xlsx(x = d, path = "tables/comorbidities_fisher.xlsx")
```



# 5. Bargraph of helmet usage

```{r}
# Data wrangling
helmet_bar_df <- as.data.frame.matrix(table(trauma_filt$alcohol.use, trauma_filt$helmet.factor))
helmet_bar_df <- tibble::rownames_to_column(helmet_bar_df, var = "alcohol")
helmet_bar_df$alcohol <- factor(helmet_bar_df$alcohol, levels = c("Not Screened", "Sober", "Impaired", "Intoxicated", "Multi-Drug"))
helmet_bar_df$percent <- (helmet_bar_df$Helmet / (helmet_bar_df$Helmet + helmet_bar_df$`No Helmet`)) * 100
helmet_bar_df$percent <- round(helmet_bar_df$percent, digits = 1)

# Barplot
helmet_bar_plot <- helmet_bar_df %>% ggplot(aes(x = alcohol, y = percent, fill = alcohol, color = "black")) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = percent), vjust = -0.5, size = 4) +
  scale_y_continuous(limits = c(0, 50), expand = c(0,0)) +
  scale_fill_manual(values = c("black", "#84888e", "#89C5D3", "#FEDBB2", "#E8ACBD")) + 
  scale_color_manual(values = "black") +
  labs(x = "BAC Intoxication Status", y = "Percent Wearing a Helmet (%)", title = "Helmet Usage") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12),
        legend.position = "none")
print(helmet_bar_plot)

# Optional Save 
#ggsave(plot = helmet_bar_plot, filename = "figures/helmet_use_barplot.pdf", width = 1400, height = 1500, dpi = 300, units = "px")
```



# 6. Dumbbell plot for comorbidities - start here

```{r}
# Function to create a comorbidity dataframe for graphing in ggplot
create_comor_df <- function(comorbid_values, comorbidity_title) {
  comor_df <- ftable(trauma_filt$alcohol.use, comorbid_values, trauma_filt$helmet.factor)
  comor_df <- data.frame(expand.grid(rev(attr(comor_df, "row.vars"))), unclass(comor_df))
  comor_df$percent <- (comor_df$X1 / (comor_df$X1 + comor_df$X2)) * 100
  colnames(comor_df) <- c("comorbidity", "alcohol", "helmet", "no_helmet", "percent")
  comor_df$comorbidity <- plyr::mapvalues(x = comor_df$comorbidity, from = c(0, 1), to = c("no", "yes"))
  comor_df$comorbidity <- factor(x = comor_df$comorbidity, levels = c("yes", "no"))
  comor_df$alcohol <- factor(x = comor_df$alcohol, levels = c("Not Screened", "Multi-Drug", "Intoxicated", "Impaired", "Sober"))
  comor_df$comorbidity_name <- comorbidity_title
  return(comor_df)
}
```

```{r}
# Calculating population counts for each comorbidity x alcohol intoxication status x helmet use group to find the interaction between addictive comorbidities and alcohol consumption in helmet use rates
comor_alc <- create_comor_df(comorbid_values = trauma_filt$CC_ALCOHOLISM, comorbidity_title = "Alcoholism")
comor_smo <- create_comor_df(comorbid_values = trauma_filt$CC_SMOKING, comorbidity_title = "Smoking")
comor_sub <- create_comor_df(comorbid_values = trauma_filt$CC_SUBSTANCEABUSE, comorbidity_title = "Substance Abuse Disorder")
#comor_men <- create_comor_df(comorbid_values = trauma_filt$CC_MENTALPERSONALITY, comorbidity_title = "Mental/Personality Disorder")
comor_df <- dplyr::bind_rows(comor_alc, comor_smo, comor_sub)  #comor_men
comor_df$comorbidity_name <- plyr::mapvalues(x = comor_df$comorbidity_name, from = c("Alcoholism", "Smoking", "Substance Abuse Disorder"), to = c("Alcohol Use Disorder", "Tobacco Use Disorder", "Substance Use Disorder"))
comor_df$comorbidity_name <- factor(comor_df$comorbidity_name, levels = c("Alcohol Use Disorder", "Tobacco Use Disorder", "Substance Use Disorder"))
comor_df$alcohol <- factor(comor_df$alcohol, levels = rev(c("Not Screened", "Sober", "Impaired", "Intoxicated", "Multi-Drug")))

# Dumbbell Plot
comor_dumb_plot <- comor_df %>% ggplot(aes(x = alcohol, y = percent)) +
  geom_line() + 
  geom_point(aes(color = comorbidity), size = 3) +
  facet_grid(cols = vars(comorbidity_name)) +
  scale_x_discrete(limits = rev) +
  scale_color_manual(values = c("#e41a1c", "#367db7")) +
  labs(x = "BAC Status", y = "Percent Helmet Wearing (%)", title = "Comorbidities in Alcohol Screening Groups", color = "Comorbidity") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 10),
        strip.background = element_blank(),
        strip.text = element_text(size = 12))
comor_dumb_plot

# Optional Save 
#ggsave(plot = comor_dumb_plot, filename = "figures/comorbidities_dumbbell.pdf", width = 3000, height = 1500, dpi = 300, units = "px")
```

```{r}
# Additional Plot for dumbbell
rate_decline_plot <- comor_df %>% ggplot(aes(x = alcohol, y = percent, group = comorbidity_status, color = comorbidity_status)) + 
  geom_line() + 
  geom_point(size = 2) + 
  scale_x_discrete(limits = rev) +
  scale_color_manual(values = c("#3fe1cf", "#367db7", "#0080ff", "#010080", "#fe0903", "#b90e0d", "#e1105d", "#800000")) +
  labs(x = "Intoxication Status", y = "Percent Helmet Wearing (%)", title = "Comorbidities in Alcohol Screening Groups", color = "Comorbidity") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10))
rate_decline_plot

# Optional Save 
#ggsave(plot = rate_decline_plot, filename = paste0(cwd, "/figures/rate_decline_plot.pdf"), width = 2200, height = 1500, dpi = 300, units = "px")
```

## 6.1 Statistics for dumbell plot - GLM

```{r}
comor_stats_df <- trauma_filt %>% dplyr::select(ALCOHOLSCREENRESULT, alcohol.use, CC_ALCOHOLISM, CC_SMOKING, CC_SUBSTANCEABUSE, CC_MENTALPERSONALITY, helmet.factor)
comor_stats_df$alcohol.use <- dplyr::recode(comor_stats_df$alcohol.use, "Not Screened" = 0, "Sober" = 1, "Impaired" = 2, "Intoxicated" = 3, "Multi-Drug" = 4)
comor_stats_df$helmet.factor <- dplyr::recode(comor_stats_df$helmet.factor, "Helmet" = 0, "No Helmet" = 1)

# Complete model with all interacting comorbidities
cc_alc_val_glm <- glm(helmet.factor ~ alcohol.use + CC_ALCOHOLISM + CC_SMOKING + CC_SUBSTANCEABUSE, family = binomial(), data = comor_stats_df)
summary(cc_alc_val_glm)
print(exp(coef(cc_alc_val_glm)))

cc_alc_val_glm <- glm(helmet.factor ~ alcohol.use * (CC_ALCOHOLISM + CC_SMOKING + CC_SUBSTANCEABUSE), family = binomial(), data = comor_stats_df)
summary(cc_alc_val_glm)
print(exp(coef(cc_alc_val_glm)))

# GLM alcoholism + alcohol use
cc_alc_glm <- glm(helmet.factor ~ alcohol.use * CC_ALCOHOLISM, family = binomial, data = comor_stats_df)
summary(cc_alc_glm)
print(exp(coef(cc_alc_glm)))

# GLM smoking + alcohol use
cc_smo_glm <- glm(helmet.factor ~ alcohol.use * CC_SMOKING, family = binomial, data = comor_stats_df)
summary(cc_smo_glm)
print(exp(coef(cc_smo_glm)))

# GLM substance use disorder + alcohol use
cc_sud_glm <- glm(helmet.factor ~ alcohol.use * CC_SUBSTANCEABUSE, family = binomial, data = comor_stats_df)
summary(cc_sud_glm)
print(exp(coef(cc_sud_glm)))
```

```{r}
# Modeling model correctness with DHARMa
simulationOutput <- simulateResiduals(fittedModel = cc_alc_val_glm)
plot(simulationOutput)

#simulationOutput <- simulateResiduals(fittedModel = cc_alc_glm)
#plot(simulationOutput)

#simulationOutput <- simulateResiduals(fittedModel = cc_smo_glm)
#plot(simulationOutput)

#simulationOutput <- simulateResiduals(fittedModel = cc_sud_glm)
#plot(simulationOutput)

#simulationOutput <- simulateResiduals(fittedModel = cc_mpd_glm)
#plot(simulationOutput)
```



# 7. Logistic Regression Analysis

```{r}
# Cleaning the trauma df for logistic regression
trauma_filt_lr <- trauma_filt %>% 
  select(sex.factor, AGEYEARS, age.factor, race.factor, ethnicity.factor, ALCOHOLSCREENRESULT, alcohol.use, helmet.factor,
         CC_ALCOHOLISM, CC_SMOKING, CC_SUBSTANCEABUSE) %>% 
  mutate(
    sex.factor = sex.factor %>% factor(levels = c("Female", "Male")) %>% ff_label("Sex"),
    ethnicity.factor = ethnicity.factor %>% factor(levels = c("Not Hispanic", "Hispanic")) %>% ff_label("Ethnicity"),
    CC_ALCOHOLISM = CC_ALCOHOLISM %>% factor() %>% fct_recode("Negative" = "0", "Positive" = "1") %>% ff_label("Alcohol Use Disorder"),
    CC_SMOKING = CC_SMOKING %>% factor() %>% fct_recode("Negative" = "0", "Positive" = "1") %>% ff_label("Tobacco Use Disorder"),
    CC_SUBSTANCEABUSE = CC_SUBSTANCEABUSE %>% factor() %>% fct_recode("Negative" = "0", "Positive" = "1") %>% ff_label("Substance Use Disorder"),
    AGEYEARS = AGEYEARS %>% ff_label("Age (years)")
    )

# Performing Logistic Regression and testing different variables in the model of predicting risk factors of helmet wearing
dependent <- "helmet.factor"
explanatory <- c("AGEYEARS", "sex.factor", "race.factor", "ethnicity.factor", "alcohol.use", 
                 "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE")
explanatory_multi <- c("AGEYEARS", "race.factor", "ethnicity.factor", "alcohol.use", 
                       "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE")
logistic_fit <- trauma_filt_lr %>% 
  finalfit(dependent, explanatory, explanatory_multi, keep_models = TRUE, metrics = TRUE)
logistic_fit1 <- logistic_fit[[1]] %>% as.data.frame()
logistic_fit2 <- logistic_fit[[2]] %>% as.character()
print(logistic_fit2)
```

```{r}
# Final logisitic regression to export
dependent <- "helmet.factor"
explanatory <- c("AGEYEARS", "sex.factor", "race.factor", "ethnicity.factor", "alcohol.use", 
                 "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE")
logistic_fit <- trauma_filt_lr %>% 
  finalfit(dependent, explanatory, metrics = TRUE)
logistic_fit1 <- logistic_fit[[1]] %>% as.data.frame()
logistic_fit2 <- logistic_fit[[2]] %>% as.character()
print(logistic_fit2)

# Exporting to excel
#write_xlsx(x = logistic_fit1, path = "tables/table5_logistic_regression.xlsx")
```

```{r}
# Odds ratio Plot
dependent <- "helmet.factor"
explanatory <- c("AGEYEARS", "sex.factor", "race.factor", "ethnicity.factor", "alcohol.use", 
                 "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE")

# Exporting a Odds ratio plot
#pdf(file = "figures/logistic_regression_odds_ratio_plot.pdf", height = 6, width = 9)
trauma_filt_lr %>% 
  or_plot(dependent, explanatory,
          remove_ref = TRUE,
          breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8),
          table_text_size = 3.5,
          title_text_size = 16)
#dev.off()
```




# 8. Supplemental Data

## 8.1 Yearly plot to get time-data for COVID

```{r}
# Function for plotting graphs
line_plotter <- function(df, x_axis, y_axis, x_label, y_label, plot_title) {
  line_plot <- df %>% 
    ggplot(aes(x = x_axis, y = y_axis)) +
    geom_line(aes(group=1), size = 1) +  
    geom_point(size = 2) +
    labs(x = x_label, y = y_label, title = plot_title) + 
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, size = 22),
        strip.background = element_blank(),
        strip.text = element_text(size = 14),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16))
  return(line_plot)
}
```

```{r}
# Getting the number of bike injuries per year
bike_injury_counts <- as.data.frame(table(trauma$Year))
colnames(bike_injury_counts) <- c("Year", "n_count")

# Yearly bike injury frequency (from python)
bike_injury_counts$ntdb_totals <- c(997970, 1043736, 1097190, 1133053, 1209097, 1232956) #2017, 2018, 2019, 2020, 2021, 2022

# Bike Injury frequency over total injuries
bike_injury_counts$bike_freq <- (bike_injury_counts$n_count / bike_injury_counts$ntdb_totals) * 100

# Plotting
bike_injury_counts_plot <- line_plotter(df = bike_injury_counts,
                                        x_axis = bike_injury_counts$Year, y_axis = bike_injury_counts$bike_freq,
                                        x_label = "Year", y_label = "Percent Pedalcylist Injuries",
                                        plot_title = "Pedalcylist Trauma Frequency")
bike_injury_counts_plot

# Optional Save 
#ggsave(plot = bike_injury_counts_plot, filename = paste0(cwd, "/figures/yearly_pedalcyclist_trauma_frequency.pdf"), width = 1600, height = 1500, dpi = 300, units = "px")
```

```{r}
# Getting the intoxication rate
intox_counts <- as.data.frame.matrix(table(trauma_filt$Year, trauma_filt$alcohol.use))
intox_counts$totals <- rowSums(intox_counts)
intox_counts$intox_freq <- ((intox_counts$Intoxicated + intox_counts$`Multi-Drug`) / intox_counts$totals) * 100  #count both intoxicated & multi-drug groups?
intox_counts <- tibble::rownames_to_column(intox_counts, var = "Year")

# Plotting
intox_freq_plot <- line_plotter(df = intox_counts,
                                x_axis = intox_counts$Year, y_axis = intox_counts$intox_freq,
                                x_label = "Year", y_label = "Percent Alcohol Intoxication",
                                plot_title = "Intoxication Rate per Year")
intox_freq_plot

# Optional Save 
#ggsave(plot = intox_freq_plot, filename = paste0(cwd, "/figures/yearly_intoxication_frequency.pdf"), width = 1600, height = 1500, dpi = 300, units = "px")
```

```{r}
# Getting the Helmet Use per year
helmet_counts <- as.data.frame.matrix(table(trauma_filt$Year, trauma_filt$helmet.factor))
helmet_counts$totals <- rowSums(helmet_counts)
helmet_counts$helmet_freq <- ((helmet_counts$Helmet) / helmet_counts$totals) * 100
helmet_counts <- tibble::rownames_to_column(helmet_counts, var = "Year")

# Plotting
helmet_freq_plot <- line_plotter(df = helmet_counts,
                                x_axis = helmet_counts$Year, y_axis = helmet_counts$helmet_freq,
                                x_label = "Year", y_label = "Percent Helmet Use",
                                plot_title = "Helmet Use per Year")
helmet_freq_plot

# Optional Save 
#ggsave(plot = helmet_freq_plot, filename = paste0(cwd, "/figures/yearly_helmet_use_frequency.pdf"), width = 1600, height = 1500, dpi = 300, units = "px")
```

```{r}
# Statistics on yearly intoxication rates
# Dependent variable of interest
dependent <- "year.factor"
# Explanatory & confounding variables
explanatory <- c("alcohol.use", "helmet.factor")
# Table
yearly_table <- trauma_filt %>% summary_factorlist(dependent, explanatory, 
                                                   p_cat = "chisq", p = TRUE)
```



## 8.2 Alcohol AIS analysis 

```{r}
# Filtering for filtered INC codes
ais_filt <- ais[ais$INC_KEY %in% trauma_filt$INC_KEY, ]
ais_filt <- ais_filt[ais_filt$AISSEVERITY != 9, ] #filtering unknown severity scores

# Glimpsing
ais_glimpse <- ff_glimpse(ais_filt)
ais_glimpse_cont <- ais_glimpse[[1]]
ais_glimpse_cat <- ais_glimpse[[2]]
```

```{r}
# Labeling AIS scores based on alcohol intoxication levels
alc_unscreened_keys <- trauma_filt %>% filter(alcohol.use == "Not Screened") %>% pull(INC_KEY)
alc_sober_keys <- trauma_filt %>% filter(alcohol.use == "Sober") %>% pull(INC_KEY)
alc_impair_keys <- trauma_filt %>% filter(alcohol.use == "Impaired") %>% pull(INC_KEY)
alc_intox_keys <- trauma_filt %>% filter(alcohol.use == "Intoxicated") %>% pull(INC_KEY)
multi_drug_keys <- trauma_filt %>% filter(alcohol.use == "Multi-Drug") %>% pull(INC_KEY)

# Labeleding AIS scores based on helmet use
helmet_yes_keys <- trauma_filt %>% filter(helmet.factor == "Helmet") %>% pull(INC_KEY)
helmet_no_keys <- trauma_filt %>% filter(helmet.factor == "No Helmet") %>% pull(INC_KEY)

ais_filt <- ais_filt %>% mutate(
  alcohol.use = case_when(
    INC_KEY %in% alc_unscreened_keys ~ "Not Screened",
    INC_KEY %in% alc_sober_keys ~ "Sober",
    INC_KEY %in% alc_impair_keys ~ "Impaired",
    INC_KEY %in% alc_intox_keys ~ "Intoxicated",
    INC_KEY %in% multi_drug_keys ~ "Multi-Drug"
  ),
  helmet.factor = case_when(
    INC_KEY %in% helmet_yes_keys ~ "Helmet",
    INC_KEY %in% helmet_no_keys ~ "No Helmet"
  )
)

# Filtering for NAs in selected columns
ais_filt <- ais_filt %>% filter_at(vars(AISPREDOT, AISSEVERITY, AISSEVERITY, ISSREGION, alcohol.use, helmet.factor), all_vars(!is.na(.)))

# Refactoring ISS region
ais_filt <- ais_filt %>% mutate(
  ISSREGION.factor = ISSREGION %>% 
    factor() %>% 
    fct_recode(
      "Head & Neck" = "1",
      "Chest" = "2",
      "Abdominal & Pelvic" = "3",
      "Extremities" = "4",
      "Face" = "5",
      "External" = "6"
  ) %>% 
    ff_label("ISS Region"),
  
  AISSEVERITY = AISSEVERITY %>% 
    factor() %>% 
    ff_label("AIS Severity Score"),
  
  alcohol.use = alcohol.use %>% 
    factor(levels = c("Not Screened", "Sober", "Impaired", "Intoxicated", "Multi-Drug")) %>% 
    ff_label("Alcohol Use"),
  
  helmet.factor = helmet.factor %>% 
    factor(levels = c("Helmet", "No Helmet")) %>% 
    ff_label("Helmet")
)
ais_filt <- ais_filt %>% mutate(
  ISSREGION = as.numeric(ISSREGION),
  AISSEVERITY = as.numeric(AISSEVERITY)
)
```

```{r}
# Filters for mild and severe AIS injury scores
#ais_filt_mild <- ais_filt[ais_filt$AISSEVERITY %in% c("1,", "2"), ]
ais_filt_severe <- ais_filt[ais_filt$AISSEVERITY > 2, ]

# Explanatory & confounding variables
explanatory <- c("ISSREGION.factor", "AISSEVERITY")

# Explanatory variable of interest
dependent <- "helmet.factor"  #alcohol.use or helmet.factor

table4.1 <- ais_filt %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE, column = TRUE,
                     add_col_totals = TRUE)
table4.1$severity <- "All"

table4.2 <- ais_filt_severe %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE,
                     add_col_totals = TRUE)
table4.2$severity <- "Severe"

table4 <- bind_rows(table4.1, table4.2)

# Optional export to excel
write_xlsx(x = table4, path = "tables/table4_ais.xlsx")
```

```{r}
# Fishers test for ISS region - all
rwf1 <- table(ais_filt$ISSREGION.factor, ais_filt$helmet.factor)
rwf1 <- rwf1[, c(2,1)]
chisq.test(rwf1)
rwf2 <- row_wise_fisher_test(rwf1, p.adjust.method = "fdr", detailed = TRUE)
```

```{r}
# Fishers test for ISS region - severe
ais_filt2 <- ais_filt[ais_filt$AISSEVERITY > 2, ]
rwf1 <- table(ais_filt2$ISSREGION.factor, ais_filt2$helmet.factor)
rwf1 <- rwf1[, c(2,1)]
chisq.test(rwf1)
rwf2 <- row_wise_fisher_test(rwf1, p.adjust.method = "fdr", detailed = TRUE)
```

```{r}
# AIS Severity scores
wilcox.test(AISSEVERITY ~ helmet.factor, data = ais_filt)

s <- ais_filt %>% dplyr::group_by(helmet.factor) %>% 
  dplyr::summarise(mean_iss = mean(AISSEVERITY))
```

```{r}
# Assessing poly-trauma: patients with more than one injury
ais_poly_counts <- ais_filt %>% group_by(INC_KEY, alcohol.use) %>% 
  dplyr::summarise(poly_trauma = n())

# Explanatory variables
explanatory <- "alcohol.use"

# Explanatory variable of interest
dependent <- "poly_trauma"

# Statistics on AIS poly-trauma counts
ais_poly_table <- ais_poly_counts %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE)

# Assessing poly-region-trauma: patients with more than one body region injury
ais_poly_region_counts <- ais_filt %>% group_by(INC_KEY, alcohol.use, ISSREGION.factor) %>% 
  dplyr::summarise(poly_trauma = n())
ais_poly_region_counts <- ais_poly_region_counts %>% group_by(INC_KEY, alcohol.use) %>% 
  dplyr::summarise(poly_trauma = n())

# Explanatory variables
explanatory <- "alcohol.use"

# Explanatory variable of interest
dependent <- "poly_trauma"

# Statistics on AIS poly-trauma counts
ais_poly_region_table <- ais_poly_region_counts %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE)


# Merging
ais_poly_table$ais_group <- "Total AIS Injuries"
ais_poly_region_table$ais_group <- "Number of AIS Body Region Injuries"
ais_poly <- rbind(ais_poly_table, ais_poly_region_table)

# Exporting
#write_csv(x = ais_poly, file = paste0(cwd, "/tables/ais_poly-injuries.csv"))
```










































