---
title: "Statistical_Analysis.Rmd"
author: "Daniel Brock"
date: "8/2/2023"
output: html_document
---

# Importing Packages and setting working directory

```{r setup, include=FALSE}
# Packages 
library(tidyverse)
library(finalfit)
library(broom)
library(ggfortify)
library(patchwork)
library(magrittr)
library(forcats)
library(survival)
library(survminer)
library(readxl)
library(writexl)
library(pheatmap)
library(zoo)

# Setting working directory
cwd <- getwd()
```

# Importing merged trauma dataset from python export

```{r}
trauma <- read_csv(file = paste0(cwd, "/TQP_Processed/trauma_merged.csv"))
comor <- read_csv(file = paste0(cwd, "/TQP_Processed/comorbidities_merged.csv"))
adverse <- read_csv(file = paste0(cwd, "/TQP_Processed/adverse_events_merged.csv"))
ais <- read_csv(file = paste0(cwd, "/TQP_Processed/ais_merged.csv"))
```

# Viewing the data and dropping important NA rows

```{r}
# Replacing quote marks in ecodes
trauma$ECODE_DESC <- gsub(pattern = "\"", replacement = "", x = trauma$ECODE_DESC)

# Merging trauma with comorbidities
trauma <- merge(trauma, comor, on = "INC_KEY")

# Merging trauma with adverse events
trauma <- merge(trauma, adverse, on = "INC_KEY")

# Glimpsing
trauma_glimpse <- ff_glimpse(trauma)
trauma_glimpse_continuous <- trauma_glimpse$Continuous
trauma_glimpse_categorical <- trauma_glimpse$Categorical
```

# Refactoring data for table 1

```{r}
# Refactoring Sex
trauma <- trauma %>% filter(SEX != 3) %>% mutate(
  sex.factor = SEX %>% 
    factor() %>% 
    fct_recode("Male" = "1", "Female" = "2") %>% 
    ff_label("Sex")
)

# Refactoring Age
trauma <- trauma %>% 
  mutate(
    age.factor = 
      AGEYEARS %>%
      cut(breaks = c(0,10,20,30,40,50,60,70,100), include.lowest = TRUE, right = FALSE) %>% 
      fct_recode(
        "0-9"   =  "[0,10)",
        "10-19" = "[10,20)",
        "20-29" = "[20,30)",
        "30-39" = "[30,40)",
        "40-49" = "[40,50)",
        "50-59" = "[50,60)",
        "60-69" = "[60,70)",
        "70+"   = "[70,100]",
      ) %>% 
      ff_label("Age (years)")
    )

# Refactoring Race 
trauma <- trauma %>% mutate(
  race.factor = case_when(
    ASIAN == 1 ~ "Asian",
    PACIFICISLANDER == 1 ~ "Pacific Islander",
    RACEOTHER == 1 ~ "Other",
    AMERICANINDIAN == 1 ~ "Native American",
    BLACK == 1 ~ "Black",
    WHITE == 1 ~ "White",
    RACE_NA == 1 ~ "NA",
    RACE_UK == 1 ~ "NA"
  )
)
trauma <- trauma %>% mutate(
  race.factor = na_if(race.factor, "NA") %>% 
    factor(levels = c("White", "Black", "Asian", "Native American", "Pacific Islander", "Other")) %>% 
    ff_label("Race")
)

# Refactoring Ethnicity
trauma <- trauma %>% mutate(
  ethnicity.factor = ETHNICITY %>% 
    factor() %>% 
    fct_recode("Hispanic" = "1", "Not Hispanic" = "2") %>% 
    ff_label("Ethnicity")
)

# Refactoring Helmet Wearing
trauma <- trauma %>% mutate(
  helmet.factor = case_when(
    PROTDEV_HELMET == 1 & PROTDEV_UK != 1 ~ "Helmet",
    PROTDEV_HELMET == 0 & PROTDEV_UK != 1 ~ "No Helmet",
    PROTDEV_UK == 1 ~ "NA"
  )
)
trauma <- trauma %>% mutate(
  helmet.factor = na_if(helmet.factor, "NA") %>% 
    factor(levels = c("Helmet", "No Helmet")) %>% 
    ff_label("Helmet Use")
)

# Refactoring hypotension
trauma <- trauma %>% mutate(
  SBP = SBP %>% ff_label("SBP mmHg")
)
trauma <- trauma %>% mutate(
  sbp.hypo = case_when(
    SBP < 90 ~ "Hypotensive",
    SBP >= 90 ~ "Not Hypotensive"
  ) %>% 
    ff_label("SBP mmHg <90")
)

# Refactoring TOTALGCS
trauma <- trauma %>% mutate(
  TOTALGCS = TOTALGCS %>% ff_label("Total GCS")
)


# Multiple drug use
trauma$total.drugs <- trauma$DRGSCR_AMPHETAMINE + trauma$DRGSCR_BARBITURATE + trauma$DRGSCR_BENZODIAZEPINES + trauma$DRGSCR_COCAINE + trauma$DRGSCR_METHAMPHETAMINE + trauma$DRGSCR_ECSTASY + trauma$DRGSCR_METHADONE + trauma$DRGSCR_OPIOID + trauma$DRGSCR_OXYCODONE + trauma$DRGSCR_PHENCYCLIDINE + trauma$DRGSCR_TRICYCLICDEPRESS + trauma$DRGSCR_CANNABINOID + trauma$DRGSCR_OTHER

# Alcohol use categories
trauma <- trauma %>% mutate(
  alcohol.use = case_when(
    ALCOHOLSCREENRESULT < 0.02 & total.drugs == 0 ~ "Sober",
    ALCOHOLSCREENRESULT >= 0.02 & ALCOHOLSCREENRESULT < 0.08 & total.drugs == 0 ~ "Impaired",
    ALCOHOLSCREENRESULT >= 0.08 & total.drugs == 0 ~ "Intoxicated"
  ) %>% 
    factor(levels = c("Sober", "Impaired", "Intoxicated")) %>% 
    ff_label("Alcohol")
)

# Refactoring Drug use
trauma <- trauma %>%
  mutate(
    drug.use = case_when(
      DRGSCR_NONE == 1 & total.drugs < 1 & ALCOHOLSCREENRESULT < 0.02 ~ "None",
      DRGSCR_AMPHETAMINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Amphetamine",
      DRGSCR_BARBITURATE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Barbiturate",
      DRGSCR_BENZODIAZEPINES == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Benzodiazepines",
      DRGSCR_COCAINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Cocaine",
      DRGSCR_METHAMPHETAMINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Methamphetamine",
      DRGSCR_ECSTASY == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Ecstasy",
      DRGSCR_METHADONE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Methadone",
      DRGSCR_OPIOID == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Opioids",
      DRGSCR_OXYCODONE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Oxycodone",
      DRGSCR_PHENCYCLIDINE == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Phencyclidine",
      DRGSCR_TRICYCLICDEPRESS == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Tricyclic Antidepressants",
      DRGSCR_CANNABINOID == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Cannabis",
      DRGSCR_OTHER == 1 & total.drugs == 1 & ALCOHOLSCREENRESULT < 0.02 ~ "Other",
      total.drugs >= 2 & ALCOHOLSCREEN < 0.02 ~ "Multiple Drugs",
      total.drugs >= 1 & ALCOHOLSCREENRESULT >= 0.02 ~ "Drugs and Alcohol",
      TRUE ~ NA_character_  # Add a default value
    ) %>% 
      factor(levels = c("None", "Cannabis", "Amphetamine", "Opioids", "Cocaine", "Benzodiazepines", "Methamphetamine", "Oxycodone", "Barbiturate", "Phencyclidine", "Tricyclic Antidepressants", "Methadone", "Ecstasy", "Drugs and Alcohol", "Other")) %>% 
      ff_label("Drugs")
  )

# Refactoring length of hospital stay
trauma <- trauma %>% mutate(
  FINALDISCHARGEDAYS = FINALDISCHARGEDAYS %>% ff_label("Length of Hospital Stay (days)")
)

# Refactoring length of ICU stay
trauma <- trauma %>% mutate(
  TOTALICULOS = TOTALICULOS %>% ff_label("Length of ICU Stay (days)")
)

# Refactoring ISS (injury severity score)
trauma <- trauma %>% mutate(
  ISS = ISS %>% ff_label("Injury Severity Score")
)

trauma <- trauma %>% mutate(
  ISS.factor = ISS %>% 
    cut(breaks = c(0,3,8,15,24,75), include.lowest = TRUE, right = TRUE) %>% 
    fct_recode(
      "Minor (0-3)" = "[0,3]",
      "Moderate (4-8)" = "(3,8]",
      "Serious (9-15)" = "(8,15]",
      "Severe (16-24)" = "(15,24]",
      "Critical (>25)" = "(24,75]"
    ) %>% 
    ff_label("ISS Categories")
)

# Refactoring Comorbidities
trauma <- trauma %>% mutate(
  CC_ALCOHOLISM = CC_ALCOHOLISM %>% ff_label("Alcoholism"),
  CC_SMOKING = CC_SMOKING %>% ff_label("Smoker"),
  CC_SUBSTANCEABUSE = CC_SUBSTANCEABUSE %>% ff_label("Substance Abuse Disorder"),
  CC_CHF = CC_CHF %>% ff_label("Congestive Heart Failure"),
  CC_RENAL = CC_RENAL %>% ff_label("End Stage Renal Disease"),
  CC_DIABETES = CC_DIABETES %>% ff_label("Diabetes"),
  CC_HYPERTENSION = CC_HYPERTENSION %>% ff_label("Hypertension"),
  CC_COPD = CC_COPD %>% ff_label("COPD"),
  CC_BLEEDING = CC_BLEEDING %>% ff_label("Bleeding Disorder"),
  CC_ANTICOAGULANT = CC_ANTICOAGULANT %>% ff_label("Anticoagulant Therapy"),
  CC_CIRRHOSIS = CC_CIRRHOSIS %>% ff_label("Cirrosis"),
  CC_CVA = CC_CVA %>% ff_label("Cerebrovascular Accident"),
  CC_MENTALPERSONALITY = CC_MENTALPERSONALITY %>% ff_label("Mental/Personality Disorder")
)

# Refactoring Adverse Events
trauma <- trauma %>% mutate(
  HC_ALCOHOLWITHDRAWAL = HC_ALCOHOLWITHDRAWAL %>% ff_label("Alcohol Withdrawal Syndrome"),
  HC_CARDARREST = HC_CARDARREST %>% ff_label("Cardiac Arrest"),
  HC_DVTHROMBOSIS = HC_DVTHROMBOSIS %>% ff_label("Deep Vein Thrombosis"),
  HC_EMBOLISM = HC_EMBOLISM %>% ff_label("Pulmonary Embolism"),
  HC_KIDNEY = HC_KIDNEY %>% ff_label("Acute Kidney Injury"),
  HC_RESPIRATORY = HC_RESPIRATORY %>% ff_label("Acute Respiratory Distress Syndrome"),
  HC_SEPSIS = HC_SEPSIS %>% ff_label("Sepsis"),
  HC_STROKECVA = HC_STROKECVA %>% ff_label("Stroke / CVA"),
  HC_VAPNEUMONIA = HC_VAPNEUMONIA %>% ff_label("Hosipital-acquired Pneumonia"),
  HC_INTUBATION = HC_INTUBATION %>% ff_label("Unplanned Intubation"),
  HC_UNPLANNEDICU = HC_UNPLANNEDICU %>% ff_label("Unplanned Admission to ICU"),
  HC_RETURNOR = HC_RETURNOR %>% ff_label("Unplanned Return to OR")
)

# Refactoring mortality using pulserate
trauma <- trauma %>% mutate(
  mortality.factor = case_when(
    PULSERATE == 0 ~ "Death",
    PULSERATE != 0 ~ "Alive"
  ) %>% 
    factor(levels = c("Death", "Alive")) %>% 
    ff_label("Mortality")
)

# Refactoring alcohol withdrawal syndrome
trauma <- trauma %>% mutate(
  alc.withdraw.factor = HC_ALCOHOLWITHDRAWAL %>% 
    factor() %>% 
    fct_recode("Positive" = "1", "None" = "0") %>% 
    ff_label("Alcohol Withdrawal Syndrome")
)

# Refactoring Year
trauma <- trauma %>% mutate(
  year.factor = Year %>% 
    factor() %>% 
    ff_label("Year")
)
```

# Filtering for NAs in important variables

```{r}
# Filtering for NAs in selected columns
trauma_filt <- trauma %>% filter_at(vars(sex.factor, age.factor, race.factor, ethnicity.factor, helmet.factor, SBP, sbp.hypo, TOTALGCS, 
                                         FINALDISCHARGEDAYS, ISS, ISS.factor, CC_ALCOHOLISM, CC_SMOKING, CC_SUBSTANCEABUSE, CC_CHF, CC_RENAL, 
                                         CC_DIABETES, CC_HYPERTENSION, CC_COPD, alc.withdraw.factor, mortality.factor), 
                                    all_vars(!is.na(.)))

# Seeing the total number of BAC screens for people with and without other drug use
bac_screens <- trauma_filt %>% mutate(
  alcohol.use.total = case_when(
    ALCOHOLSCREENRESULT < 0.02 ~ "Sober",
    ALCOHOLSCREENRESULT >= 0.02 & ALCOHOLSCREENRESULT < 0.08 ~ "Impaired",
    ALCOHOLSCREENRESULT >= 0.08 ~ "Intoxicated"
  ),
  alcohol.use.clean = case_when(
    ALCOHOLSCREENRESULT < 0.02 & total.drugs == 0 ~ "Sober",
    ALCOHOLSCREENRESULT >= 0.02 & ALCOHOLSCREENRESULT < 0.08 & total.drugs == 0 ~ "Impaired",
    ALCOHOLSCREENRESULT >= 0.08 & total.drugs == 0 ~ "Intoxicated"
  )
)
bac_screens_total <- sum(!is.na(bac_screens$alcohol.use.total))
bac_screens_clean <- sum(!is.na(bac_screens$alcohol.use.clean))
print(paste0("Number of total BAC screens in patients with known info: ", bac_screens_total))
print(paste0("Number of drug-free BAC screens in patients with known info: ", bac_screens_clean))

# Filtering for only patients with BAC screens with NO other drug use (clean)
keys_to_keep <- bac_screens[!is.na(bac_screens$alcohol.use.clean), ] %>% pull(INC_KEY)
trauma_filt <- trauma_filt %>% filter(INC_KEY %in% keys_to_keep)
```

# Making a demographics table for alcohol use

```{r}
# Explanatory & confounding variables
explanatory <- c("sex.factor", "age.factor", "race.factor", "ethnicity.factor")

# Dependent variable of interest
dependent <- "alcohol.use"

table1 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE)

# Optional export to excel
#write_csv(x = table1, file = paste0(cwd, "/tables/table1_demographics.csv"))
```

# Making a demographics table comparing alcohol use medical outcomes

```{r}
# Explanatory & confounding variables
explanatory <- c("helmet.factor", 
                 "mortality.factor", "FINALDISCHARGEDAYS", "TOTALICULOS", 
                 "ISS", "ISS.factor", "TOTALGCS", "SBP", "sbp.hypo", 
                 "HC_ALCOHOLWITHDRAWAL", "HC_CARDARREST", "HC_DVTHROMBOSIS", "HC_EMBOLISM", "HC_KIDNEY", "HC_RESPIRATORY", "HC_SEPSIS", "HC_STROKECVA", "HC_VAPNEUMONIA", "HC_INTUBATION", "HC_UNPLANNEDICU", "HC_RETURNOR",
                 "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE", "CC_MENTALPERSONALITY", "CC_CHF", "CC_RENAL", "CC_DIABETES", "CC_HYPERTENSION", "CC_COPD", "CC_BLEEDING", "CC_ANTICOAGULANT", "CC_CIRRHOSIS", "CC_CVA")

# Dependent variable of interest
dependent <- "alcohol.use"

table2 <- trauma_filt %>% 
  summary_factorlist(dependent, explanatory,
                     cont = "median", p_cont_para = "kruskal.test", p_cat = "chisq", p = TRUE,
                     add_row_totals = FALSE)

# Optional export to excel
#write_csv(x = table2, path = paste0(cwd, "/tables/table2_outcomes.csv"))
```

# Bargraph of helmet use

```{r}
# Data wrangling
helmet_bar_df <- table2[table2$label == "Helmet Use", ] %>% pivot_longer(cols = c("Sober", "Impaired", "Intoxicated"), names_to = "alcohol", values_to = "percent")
helmet_bar_df$percent2 <- str_extract(helmet_bar_df$percent, "\\([^()]*\\)")
helmet_bar_df$percent2 <- gsub(pattern = "\\(", replacement = "", x = helmet_bar_df$percent2)
helmet_bar_df$percent2 <- gsub(pattern = "\\)", replacement = "", x = helmet_bar_df$percent2)
helmet_bar_df$percent2 <- as.numeric(helmet_bar_df$percent2)
helmet_bar_df$alcohol <- factor(helmet_bar_df$alcohol, levels = c("Sober", "Impaired", "Intoxicated"))

# Barplot
helmet_bar_plot <- helmet_bar_df %>% ggplot(aes(x = alcohol, y = percent2, fill = alcohol)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = percent2), vjust = -0.5) +
  scale_y_continuous(limits = c(0, 50), expand = c(0,0)) + 
  scale_fill_manual(values = c("#89C5D3", "#FEDBB2", "#E8ACBD")) + 
  labs(x = "BAC Intoxication Status", y = "Percent Wearing a Helmet (%)", title = "Helmet Usage in Alcohol Screening Groups") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "none")
helmet_bar_plot

# Optional Save 
#ggsave(plot = helmet_bar_plot, filename = paste0(cwd, "/figures/helmet_use_barplot.pdf"), width = 1500, height = 1500, dpi = 300, units = "px")
```

# Bargraph of comorbidities - start here

```{r}
# Data wrangling
comor_df <- replace(x = table2, table2=="", NA)
comor_df <- na.locf(na.locf(comor_df), fromLast = TRUE)
comorbidities <- c("Alcoholism", "Smoker", "Substance Abuse Disorder", "Mental/Personality Disorder")
comor_df <- comor_df[comor_df$label %in% comorbidities & comor_df$levels == 1, ]
colnames(comor_df) <- c("Comorbidities", "levels", "Sober", "Impaired", "Intoxicated", "p")
comor_df$Comorbidities <- gsub(comor_df$Comorbidities, pattern = "Substance Abuse Disorder", replacement = "Substance Abuse")
comor_df$Comorbidities <- gsub(comor_df$Comorbidities, pattern = "Mental/Personality Disorder", replacement = "Mental Disorder")
comor_df <- comor_df %>% pivot_longer(cols = c("Sober", "Impaired", "Intoxicated"), names_to = "Intoxication_Status", values_to = "Percent")
comor_df$Comorbidities <- factor(x = comor_df$Comorbidities, levels = c("Alcoholism", "Smoker", "Substance Abuse", "Mental Disorder"))
comor_df$Intoxication_Status <- factor(x = comor_df$Intoxication_Status, levels = c("Sober", "Impaired", "Intoxicated"))
comor_df$Percent <- str_extract(comor_df$Percent, "\\([^()]*\\)")
comor_df$Percent <- gsub(pattern = "\\(", replacement = "", x = comor_df$Percent)
comor_df$Percent <- gsub(pattern = "\\)", replacement = "", x = comor_df$Percent)
comor_df$Percent <- as.numeric(comor_df$Percent)

# Grouped Barplot
comor_bar_plot <- comor_df %>% ggplot(aes(x = Comorbidities, y = Percent, group = Intoxication_Status, fill = Intoxication_Status)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  scale_y_continuous(limits = c(0, 35), expand = c(0,0)) + 
  scale_fill_manual(values = c("#89C5D3", "#FEDBB2", "#E8ACBD")) + 
  labs(x = "Comorbidities", y = "Percent of Intoxication Groups with Comorbidity (%)", title = "Comorbidities in Alcohol Screening Groups", fill = "Intoxication Status") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12))
comor_bar_plot

# Optional Save 
#ggsave(plot = comor_bar_plot, filename = paste0(cwd, "/figures/comorbidities.pdf"), width = 2500, height = 1500, dpi = 300, units = "px")
```

# Yearly plot to get time-data for COVID

```{r}
# Function for plotting graphs
line_plotter <- function(df, x_axis, y_axis, x_label, y_label, plot_title) {
  line_plot <- df %>% 
    ggplot(aes(x = x_axis, y = y_axis)) +
    geom_line(size = 1) +  
    geom_point(size = 2) +
    labs(x = x_label, y = y_label, title = plot_title) + 
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, size = 22),
        strip.background = element_blank(),
        strip.text = element_text(size = 14),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16))
  return(line_plot)
}
```

```{r}
# Getting the number of bike injuries per year
bike_injury_counts <- trauma %>% group_by(Year) %>% 
  summarise(n_count = n())

# Yearly bike injury frequency (from python)
bike_injury_counts$ntdb_totals <- c(997970, 1043736, 1097190, 1133053, 1209097, 1232956) #2017, 2018, 2019, 2020, 2021, 2022

# Bike Injury frequency over total injuries
bike_injury_counts$bike_freq <- (bike_injury_counts$n_count / bike_injury_counts$ntdb_totals) * 100

# Plotting
bike_injury_counts_plot <- line_plotter(df = bike_injury_counts,
                                        x_axis = bike_injury_counts$Year, y_axis = bike_injury_counts$bike_freq,
                                        x_label = "Year", y_label = "Percent Pedalcylist Injuries",
                                        plot_title = "Pedalcylist Trauma Frequency")
bike_injury_counts_plot

# Optional Save 
#ggsave(plot = bike_injury_counts_plot, filename = paste0(cwd, "/figures/yearly_pedalcyclist_trauma_frequency.pdf"), width = 1600, height = 1500, dpi = 300, units = "px")
```

```{r}
# Getting the intoxication rate
intox_counts <- trauma_filt %>% group_by(Year, alcohol.use) %>% 
  summarise(count_intox = n()) %>% 
  filter(alcohol.use == "Intoxicated")
year_counts <- trauma_filt %>% group_by(Year) %>% 
  summarise(count_total = n())
intox_counts <- merge(intox_counts, year_counts, by = "Year")
intox_counts$intox_freq <- (intox_counts$count_intox / intox_counts$count_total) * 100

# Plotting
intox_freq_plot <- line_plotter(df = intox_counts,
                                x_axis = intox_counts$Year, y_axis = intox_counts$intox_freq,
                                x_label = "Year", y_label = "Percent Alcohol Intoxication",
                                plot_title = "Intoxication Rate per Year")
intox_freq_plot

# Optional Save 
#ggsave(plot = intox_freq_plot, filename = paste0(cwd, "/figures/yearly_intoxication_frequency.pdf"), width = 1600, height = 1500, dpi = 300, units = "px")
```

```{r}
# Getting the Helmet Use per year
helmet_counts <- trauma_filt %>% group_by(Year, helmet.factor) %>% 
  summarise(count_helmet = n()) %>% 
  filter(helmet.factor == "Helmet")
year_counts <- trauma_filt %>% group_by(Year) %>% 
  summarise(count_total = n())
helmet_counts <- merge(helmet_counts, year_counts, by = "Year")
helmet_counts$intox_freq <- (helmet_counts$count_helmet / helmet_counts$count_total) * 100

# Plotting
helmet_freq_plot <- line_plotter(df = helmet_counts,
                                x_axis = helmet_counts$Year, y_axis = helmet_counts$intox_freq,
                                x_label = "Year", y_label = "Percent Helmet Use",
                                plot_title = "Helmet Use per Year")
helmet_freq_plot

# Optional Save 
#ggsave(plot = helmet_freq_plot, filename = paste0(cwd, "/figures/yearly_helmet_use_frequency.pdf"), width = 1600, height = 1500, dpi = 300, units = "px")
```

```{r}
# Statistics on yearly intoxication rates
# Dependent variable of interest
dependent <- "year.factor"
# Explanatory & confounding variables
explanatory <- c("alcohol.use", "helmet.factor")
# Table
yearly_table <- trauma_filt %>% summary_factorlist(dependent, explanatory, 
                                                   p_cat = "chisq", p = TRUE)
```

# Alcohol AIS analysis 

```{r}
# Filtering for filtered INC codes
ais_filt <- ais[ais$INC_KEY %in% trauma_filt$INC_KEY, ]
ais_filt <- ais_filt[ais_filt$AISSEVERITY != 9, ] #filtering unknown severity scores

# Glimpsing
ais_glimpse <- ff_glimpse(ais_filt)
ais_glimpse_cont <- ais_glimpse[[1]]
ais_glimpse_cat <- ais_glimpse[[2]]
```

```{r}
# Labeling AIS scores based on alcohol intoxication levels
alc_sober_keys <- trauma_filt %>% filter(alcohol.use == "Sober") %>% pull(INC_KEY)
alc_impair_keys <- trauma_filt %>% filter(alcohol.use == "Impaired") %>% pull(INC_KEY)
alc_intox_keys <- trauma_filt %>% filter(alcohol.use == "Intoxicated") %>% pull(INC_KEY)

ais_filt <- ais_filt %>% mutate(
  alcohol.use = case_when(
    INC_KEY %in% alc_sober_keys ~ "Sober",
    INC_KEY %in% alc_impair_keys ~ "Impaired",
    INC_KEY %in% alc_intox_keys ~ "Intoxicated"
  )
)

# Filtering for NAs in selected columns
ais_filt <- ais_filt %>% filter_at(vars(AISPREDOT, AISSEVERITY, AISSEVERITY, ISSREGION, alcohol.use), all_vars(!is.na(.)))

# Refactoring ISS region
ais_filt <- ais_filt %>% mutate(
  ISSREGION.factor = ISSREGION %>% 
    factor() %>% 
    fct_recode(
      "Head & Neck" = "1",
      "Chest" = "2",
      "Abdominal & Pelvic" = "3",
      "Extremities" = "4",
      "Face" = "5",
      "External" = "6"
  ) %>% 
    ff_label("ISS Region"),
  
  AISSEVERITY = AISSEVERITY %>% 
    factor() %>% 
    ff_label("AIS Severity Score"),
  
  alcohol.use = alcohol.use %>% 
    factor(levels = c("Sober", "Impaired", "Intoxicated")) %>% 
    ff_label("Alcohol Use")
)
ais_filt <- ais_filt %>% mutate(
  ISSREGION = as.numeric(ISSREGION),
  AISSEVERITY = as.numeric(AISSEVERITY)
)
```

```{r}
# Filters for mild and severe AIS injury scores
ais_filt_mild <- ais_filt[ais_filt$AISSEVERITY %in% c("1,", "2"), ]
ais_filt_severe <- ais_filt[ais_filt$AISSEVERITY %in% c("3,", "4,", "5", "6"), ]

# Explanatory & confounding variables
explanatory <- c("ISSREGION.factor", "AISSEVERITY")

# Explanatory variable of interest
dependent <- "alcohol.use"

table3.1 <- ais_filt %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE, column = FALSE)
table3.1$severity <- "All"

table3.2 <- ais_filt_severe %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE)
table3.2$severity <- "Severe"

table3 <- bind_rows(table3.1, table3.2)

# Optional export to excel
#write_csv(x = table3, file = paste0(cwd, "/tables/table3_ais.csv"))
```

```{r}
# Assessing poly-trauma: patients with more than one injury
ais_poly_counts <- ais_filt %>% group_by(INC_KEY, alcohol.use) %>% 
  summarise(poly_trauma = n())

# Explanatory variables
explanatory <- "alcohol.use"

# Explanatory variable of interest
dependent <- "poly_trauma"

# Statistics on AIS poly-trauma counts
ais_poly_table <- ais_poly_counts %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE)

# Assessing poly-region-trauma: patients with more than one body region injury
ais_poly_region_counts <- ais_filt %>% group_by(INC_KEY, alcohol.use, ISSREGION.factor) %>% 
  summarise(poly_trauma = n())
ais_poly_region_counts <- ais_poly_region_counts %>% group_by(INC_KEY, alcohol.use) %>% 
  summarise(poly_trauma = n())

# Explanatory variables
explanatory <- "alcohol.use"

# Explanatory variable of interest
dependent <- "poly_trauma"

# Statistics on AIS poly-trauma counts
ais_poly_region_table <- ais_poly_region_counts %>% 
  summary_factorlist(dependent, explanatory, 
                     cont = "mean", p_cont_para = "aov", p_cat = "chisq", p = TRUE)


# Merging
ais_poly_table$ais_group <- "Total AIS Injuries"
ais_poly_region_table$ais_group <- "Number of AIS Body Region Injuries"
ais_poly <- rbind(ais_poly_table, ais_poly_region_table)

# Exporting
#write_csv(x = ais_poly, file = paste0(cwd, "/tables/ais_poly-injuries.csv"))
```

# Logistic Regression Analysis

```{r}
# Cleaning the trauma df for logistic regression
trauma_filt_lr <- trauma_filt %>% 
  select(sex.factor, AGEYEARS, age.factor, race.factor, ethnicity.factor, ALCOHOLSCREENRESULT, alcohol.use, helmet.factor,
         CC_ALCOHOLISM, CC_SMOKING, CC_SUBSTANCEABUSE, CC_MENTALPERSONALITY) %>% 
  mutate(
    sex.factor = sex.factor %>% factor(levels = c("Female", "Male")) %>% ff_label("Sex"),
    ethnicity.factor = ethnicity.factor %>% factor(levels = c("Not Hispanic", "Hispanic")) %>% ff_label("Ethnicity"),
    CC_ALCOHOLISM = CC_ALCOHOLISM %>% factor() %>% fct_recode("Negative" = "0", "Positive" = "1") %>% ff_label("Alcoholism"),
    CC_SMOKING = CC_SMOKING %>% factor() %>% fct_recode("Negative" = "0", "Positive" = "1") %>% ff_label("Smoking"),
    CC_SUBSTANCEABUSE = CC_SUBSTANCEABUSE %>% factor() %>% fct_recode("Negative" = "0", "Positive" = "1") %>% ff_label("Substance Abuse"),
    CC_MENTALPERSONALITY = CC_MENTALPERSONALITY %>% factor() %>% fct_recode("Negative" = "0", "Positive" = "1") %>% ff_label("Mental/Personality Disorder")
    )

# Performing Logistic Regression and testing different variables in the model of predicting risk factors of helmet wearing
dependent <- "helmet.factor"
explanatory <- c("sex.factor", "race.factor", "ethnicity.factor", "alcohol.use", 
                 "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE", "CC_MENTALPERSONALITY")
explanatory_multi <- c("race.factor", "ethnicity.factor", "alcohol.use", 
                       "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE", "CC_MENTALPERSONALITY")
logistic_fit <- trauma_filt_lr %>% 
  finalfit(dependent, explanatory, explanatory_multi, keep_models = TRUE, metrics = TRUE)
logistic_fit1 <- logistic_fit[[1]] %>% as.data.frame()
logistic_fit2 <- logistic_fit[[2]] %>% as.character()
print(logistic_fit2)
```

```{r}
# Final logisitic regression to export
dependent <- "helmet.factor"
explanatory <- c("sex.factor", "race.factor", "ethnicity.factor", "alcohol.use", 
                 "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE", "CC_MENTALPERSONALITY")
logistic_fit <- trauma_filt_lr %>% 
  finalfit(dependent, explanatory, metrics = TRUE)
logistic_fit1 <- logistic_fit[[1]] %>% as.data.frame()
logistic_fit2 <- logistic_fit[[2]] %>% as.character()
print(logistic_fit2)

# Exporting to excel
#write_csv(x = logistic_fit1, file = paste0(cwd, "/tables/table4_logistic_regression.csv"))
```

```{r}
# Odds ratio Plot
dependent <- "helmet.factor"
explanatory <- c("sex.factor", "race.factor", "ethnicity.factor", "alcohol.use", 
                 "CC_ALCOHOLISM", "CC_SMOKING", "CC_SUBSTANCEABUSE", "CC_MENTALPERSONALITY")

# Exporting a Odds ratio plot
#pdf(file = paste0(cwd, "/figures/logistic_regression_odds_ratio_plot.pdf"), height = 6, width = 9)
trauma_filt_lr %>% 
  or_plot(dependent, explanatory,
          remove_ref = TRUE,
          breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8),
          table_text_size = 3.5,
          title_text_size = 16)
#dev.off()
```









































